Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-06-10 man/ecosystems/ms_ecosystems.Rmd
update text

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a). The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 4a and S3a). Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a). The strongest reduction in radial growth occurred in the 1995 drought (the worst drought spell in our climatic record, Table S3) for all sites. Tree-growth reduction followed a moisture gradient. The southern sites (CA-High and CA-Low) showed less tree-growth reduction than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site (CA-High).

The response of tree growth to water availability was greater than to temperatures. Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure S6a). Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure S6a).

@@

### Resilience to drought events at the ecosystem and individual-tree levels
Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance. Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and recovery, significant for all oak populations (Figure 5a). A similar pattern was found for resilience but proved significant only for SJ. Importantly, non-significant patterns resulted when we excluded 1995, except for recovery in SJ (Figure S5). The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4b).

During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1). The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years. For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event. For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).

The resilience metrics calculated varied significantly between sites, except for resilience of tree growth (p = 0.534; Table S1), which was similar among the three sites. The two southern populations showed lower recovery values than did the northern site both for greenness and tree growth, but resistance and resilience values were significantly higher for the southern site (Table S2).

## Discussion

By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Quercus pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades. Our results indicated that these relict oak populations driven by historical land use are resilient to climate change at their present rear edge. However, resistance, resilience, and forest recovery to extreme drought events are strongly influenced by mountain exposure, local environmental conditions, and management legacies. This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance varies markedly along the rear edge under climate change.

### Land-use legacies shape sensitivity to climate change of forests and the present rear edge
The review of historical documents revealed that forest clearings, firewood removal, charcoal making, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species approaches 90% of the cover at medium and low elevations [@JimenezOlivencia2015]. Together with the analysis of the disturbance chronologies, the observed notable differences in stand structure, tree size, and age suggest different forest histories and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland). On the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an elevational gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivencia2015]. In addition, other activities such as mining must have altered the forest structure, *e.g.* the SJ site has many small mines and quarries that were exploited intermittently throughout history. The release growth event discerned for the 1940s concurs with a period of maximum mining activity in this area (1925 to 1957), during which timber use increased for mine tunnels and furnaces, these also requiring large amounts of firewood to melt the mineral (Table S4). This heavy exploitation of the neighboring forest resources must have affected a significant part of this oak woodland, as shown by growth of the remnant trees at the northern site (Figure S2b).

On the other hand, woodlands on the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks grow [@JimenezOlivencia2015]. Firewood, charcoal, and acorns have been intensively exploited at the southern sites, until at least the mid-20th century, when these activities sharply declined due mainly to rural abandonment and the use of gas and fossil fuels [@Valbuena2013]. At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management type which has been applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].

The other release event observed for the SJ site during the period 1995-2000 was lower than during 1940, but also affected most trees (Figures 4, S2b). No records of forest practices in this area over the last 30 years have been found [@Bonet2016obsnev_forest], and no logging has been recorded during the period 1995 - 2000 (F.J. Cano-Manuel *personal communication*). Therefore this release might be related to natural drought-induced mortality after 1995, as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree growth for this site show high sensitivity to water availability [@GeaIzquierdo2014].

@@

### Sensitivity (resistance) of relict oaks to recent drought events
Severe drought negatively affects both primary and secondary growth of *Quercus pyrenaica* forests. This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by the consistent radial-growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Quercus pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014]. Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].

Greenness proved less sensitive to drought than did tree growth, particularly for drier sites. These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013; @Fatichi2014]. Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].

Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations. For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014]. It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4). This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and overall both are considered to constitute the rear edge for the species.

@@

### Relict oaks show high resilience to recent drought events and long-term climatic variability 
Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Quercus pyrenaica* for the last 16 years. This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014]. This could be related to a non-linear positive effect of warming for the species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014]. Importantly, for rear edges threatened by climate change, negative growth trends would have been expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].

Although the 2012 drought event was more severe and intense than that of 2005 (Table S3), resilience values for greenness and tree growth were greater for 2012. This could be due to the different timing of the two droughts. The 2012 event was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought. The latter matched the period of maximum growth for oak forests in late spring (Figure S4). This highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Camarero2015b; @Huang2018]. For tree growth, the highest values of resilience were found for the two most severe events (1995 and 1999; Table S3) and tree-growth resilience was positively related to drought severity (Figure 5).

The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak. Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017]. @MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations. Our results highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our forest. All this, together with the characteristic high resprouting ability of the species, would suggest a long-term persistence of those populations [@Bellingham2000]. It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the trees express resilience at the species level or whether we would find vulnerability if we analyzed seedling regeneration, as in other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].

@@

In summary, two main results stand out from this research. First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019]. The current niche is a result of land-use changes, which further complicate the definition of potential rear edges. The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014].

Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies. In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study. Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the geographical rear edge studied. This suggests that the rear edge therefore needs to be redefined [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the ecological (limiting) rear edge of species.

Overall, our results show that management history influences tree growth and resilience to climate change of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017] which, hence, will also strongly determine the current geographical distribution of marginal stands, regardless of the potential extent of ecological marginality of species.

@@

## Concluding comments
The ecological and geographical rear edges did not appear to fully match in our study. Severe drought events provoke major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the relict, rear-edge forest studied. However, we found no negative growth trends despite our expectation of vulnerability to climate change for these relict stands. Furthermore, we detected positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a steep positive trend of secondary growth at the tree level at the wettest site along the climatic gradient analyzed. The trees exhibited high resilience values in response to drought, particularly in the long-term scale. These findings are consistent with other results showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*. The differences found in tree growth and resilience to drought between sites close together show that responses to drought were site dependent and can drastically vary even in very narrow spatial gradients (*i.e.* following ecological thresholds). This is particularly relevant for rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia. The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense, our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with a high sensitivity to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these singular forests offer.

>>>>

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite stand competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a). The growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4a and S3a). Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the wetter and colder high elevation site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a). The strongest reduction in radial growth occurred in response to the 1995 drought (the worst drought spell in our climatic record, Table S3) for all sites. Tree-growth reductions in response to drought followed a moisture gradient. Tree-growth reductions in response to the studied drought events were lower in the southern sites (CA-High and CA-Low) than in the northern site (SJ), especially for 2005 and 2012 (Figure S2a). The weakest growth reductions were found in trees from the wettest site (CA-High).

The response of tree growth to water availability was greater than to temperatures. Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure S6a). Nevertheless, there were differences between populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure S6a).

@@

### Resilience to drought events at the forest stand and individual-tree levels
Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance. During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1). The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years. For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event. For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).

The recovery and resistance for greenness and growth varied significantly between sites. Resilience calculated for greenness also differed between sites but not for tree growth (p = 0.534; Table S1). The two southern populations showed lower recovery values than did the northern site both for greenness and tree growth, but resistance and resilience values were significantly higher for the southern site (Table S2).

Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and recovery, significant for all oak populations (Figure 5a). A similar pattern was found for resilience but proved significant only for SJ. Importantly, non-significant patterns resulted when we excluded 1995, except for recovery in SJ (Figure S5). The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4b).

## Discussion
///////
By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Quercus pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades. 
//////

Our results indicate that these relict oak populations driven by historical land use have been resilient to climate change at their present rear edge. However, resistance, resilience, and forest recovery to extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies. This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance can vary markedly along the rear edge under climate change

### High sensitivity of oaks to recent drought events
Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by the consistent radial-growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].

Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations of Sierra Nevada. The northern site showed higher browning intensity than the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014]. Greenness proved less sensitive to drought than did tree growth, particularly for drier sites. These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013; @Fatichi2014]. Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018]. ~~It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]~~. On the other hand, trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4). This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and overall both are considered to constitute the rear edge for the species.

As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014]. Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017]. In fact, growth of some tree species in their rear edge (*e.g.* Abies alba) have also shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrez2019EvidenceThat]; others species respond simultaneously to both temperature and moisture related variables [*e.g.* *Fagus sylvatica*, @Dorado2017; @Dorado2017c; *Pinus nigra* subsp. *salzmanii*, @SanchezSalguero2012]. Even more, other species be more dependent to temperatures [*e.g.* *Pinus sylvestris*, @Herrero2013]. This diversity of responses by tree species to the two main components of climate change (*i.e.* increase of temperature and rainfall reduction) highlihts that the vulnerability to climate is not consistently expressed within the rear edge, evidencing that geographically marginal forests are not necessarily climatically marginal [see @DoradoLinanetal2019GeographicalAdaptation and references therein]. ~~A negative effect of temperature expressing a limiting effect of high vapor pressure deficit and high potential evapotranspiration is to be expected at drought-limited rear-edges. In any case, this proves that the vulnerability to climate is not consistently expressed within the rear edge, evidencing that geographically marginal forests are not necessarily climatically marginal~~

~~Esta diversidad de respuestas de las diferentes especies forestales a los dos componentes principales del cambio climático (incremento de temperatura vs reducción de precipitación) dificultan aún mas la posibilidad de encontrar un patrón general de ajustes finos entre limite ecológico y límite geográfico (see Dorado-Liñan and others 2019 and references therein).~~

### Land-use legacies shape forest sensitivity to climate change of forests at the present rear edge
The review of historical documents revealed that forest clearings, firewood removal, charcoal making, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species would approach 90% in tree cover at medium and low elevations [@JimenezOlivencia2015]. Together with the analysis of the disturbance chronologies, the observed notable differences in stand structure, tree size, and age suggest different forest histories and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland). On the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an elevational gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivencia2015]. In addition, other activities such as mining must have altered the forest structure, *e.g.* the SJ site has many small mines and quarries that were exploited intermittently throughout history. The release growth event discerned for the 1940s concurs with a period of maximum mining activity in this area (1925 to 1957), during which timber use increased for mine tunnels and furnaces, these also requiring large amounts of firewood to melt the mineral (Table S4). This heavy exploitation of the neighboring forest resources must have affected a significant part of this oak woodland, as shown by growth of the remnant trees at the northern site (Figure S2b).

On the other hand, woodlands on the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks grow [@JimenezOlivencia2015]. Firewood, charcoal, and acorns have been intensively exploited at the southern sites, until at least the mid-20th century, when these activities sharply declined due mainly to rural abandonment and the use of gas and fossil fuels [@Valbuena2013]. At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management practice often applied in the past in many Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].

The other release event observed for the SJ site during the period 1995-2000 was lower than during 1940, but also affected most trees (Figures 4b, S2b). No records of forest practices in this area over the last 30 years have been found [@Bonet2016obsnev_forest], and no logging has been recorded during the period 1995 - 2000 (F.J. Cano-Manuel *personal communication*). Therefore this release might be related to natural drought-induced mortality after 1995, as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].

### Relict oaks show high resilience to recent drought events and long-term climatic variability 
Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years. This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4a). Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014]. This could be related to a non-linear positive effect of warming for species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014]. Importantly, for rear edges threatened by climate change, negative growth trends were expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].

Although the 2012 drought event was more severe and intense than that of 2005 (Table S3), resilience values for greenness and tree growth were greater for 2012. This could be due to the different timing of the two droughts. The 2012 event was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought. The latter matched the period of maximum growth for oak forests in late spring (Figure S4). This would highlight the importance of the timing of the drought as a key factor determining tree recovery after drought [@Camarero2015b; @Huang2018]. For tree growth, the highest values of resilience were found for the two most severe events (1995 and 1999; Table S3) and tree-growth resilience was positively related to drought severity (Figure 5a).

The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak species. Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017]. In our case, this can be explained by the fact that the current geographical rear-edge does not match with the potential ecologically rear-edge for the species because this has been determined mostly by human use. @MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations. Our results highlight the ample small-scale variability at the ecological boundary and need to better define and delimit the rear-edges. All this, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@Bellingham2000]. It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the high resilience observed is expressed at the species level or only in adult trees. The rear-edge might be different for different ontogenic stages. It is important to assess whether there is vulnerability if we analyzed seedling regeneration and recruitment, as in other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].

In summary, two main results stand out from this research. First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019]. The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014]. Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies. In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study. Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the whole geographical rear edge studied. This suggests that the rear edge therefore needs to be redefined in space but also in time [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the potential ecological (limiting) rear-edge of species. The rear-edge concept should also consider historical aspects in addition to the geographic, climatic and genetic dimensions [@VilaCabrera2019], particularly in areas with a long-history of anthropic management, such as Mediterranean mountains. Therefore, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate‐driven population decline at rear edges [@VilaCabrera2019]. Furthermore, even the geographical dimension of the rear edge is not a continuous, uniform space of ecological conditions in mountain landscapes. In other words: in mountains, heterogeneity of ecological conditions at fine scales is the rule, enabling the existence of microrefugia and enhancing species persistence [@Olaldeetal2002WhiteOaksa; @SerraDiazetal2015DisturbanceClimate]. In this respect, our results showed that resilience of oak stands to drought events was not spatially homogeneous across the mountain range, due to differences in ecological conditions and/or past-management legacies. This is particularly relevant for rear-edge populations where topographic and biophysical variability can facilitate the existence of refugia.

## Concluding comments
The ecological and geographical rear edges did not appear to fully match in our study sistem. Severe drought events provoked major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the a priori relict, rear-edge forest studied. Nevertheless, contrary to our expectations, the trees exhibited high resilience in response to drought, particularly in the long-term. The differences found in tree growth and resilience to drought between close neighboring sites show that responses to drought are site dependent and can drastically vary in very narrow spatial gradients. This is particularly relevant to refine the real extension of rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia. The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents. In consequence, the geographical and ecological rear-edges for specific species do not necessarily meet after centuries or millennia of human impacts on the real world landscapes. This is relevant for tree species with a high sensitivity to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these forests offer.

9999999999999999
OLD

@@

@@

@@

@@
